Open source landscape ecology tools \thanks{Grants or other notes about the article that should go on the front page should be placed here. General acknowledgments should be placed at the end of the article.}

### 2020-07-20 paper/paper.Rmd
Replying to some comments (either by comments or changes in the text)

---
title: Open source landscape ecology tools
# subtitle: Do you have a subtitle? If so, write it here
# titlerunning: Open source landscape ecology tools
# authorrunning: Short form of author list if too long for running head
thanks: | 
    Grants or other notes about the article that should go on the front 
    page should be placed here. General acknowledgments should be placed at the
    end of the article.
authors: 
- name: Maximillian H.K. Hesselbarth
  address: Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
  email: maximilian.hesselbarth@uni-goettingen.de
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Author 3
  address: Adress author 3
  email: E-Mail author 3
- name: ...
  address: ...
  email: ...
keywords:
- keyword 1
- keyword 2
- keyword 3
- keyword 4
- ...
abstract: |
  max. (200 words) Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: | 
  \usepackage{caption}
  \usepackage{ltablex}
  \usepackage{lscape}
  \usepackage{multirow}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
editor_options: 
  chunk_output_type: console
---

@@

<!-- MH: Do we need some more basic introduction to landscape ecology? -->

<!-- JN:  We still can slightly update the landscape ecology introduction, but I think that the lenght of this text is good. Three paragraphs seems enough, as our goal is not to explain what is the landscape ecology, but what are the tools used in this field. -->

@@

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}. 
Open-source software includes all software which is released under a license that allows to freely use, modify and distribute the software \cite{St.Laurent2008}.
Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
<!--probably we should also mention the cost here-->
Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology \cite{Prlic2012,Powers2019}. 
Lastly, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} and to customize existing software to their specific needs \cite{Steiniger2009}.

One example of a successful open-source project is the *R* programming language and its *Comprehensive R Archive Network* (CRAN) for extensions (also called packages) \cite{RCoreTeam2019}. 
Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}. 
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006}.
This is documented by the growing *CRAN Task View: Analysis of Spatial Data* \cite{Bivand2019a} and *CRAN Task View: Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} that currently list almost 300 packages in total.
<!-- we should add that the total number of spatial-related packages is, in fact, larger, as the task views, are manual lists, updated just by a few people -->
The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.

Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on tools implemented in the *R* programming.
<!-- probably we should state why we are focusing on R only -->
<!-- some answers: "vastness of spatial R packages", ... -->
Thus, in the first part we are give a comprehensive overview over existing *R* packages for landscape ecology. 
In the second part, we identify topics for which *R* packages are currently missing to the best of our knowledge. 
For more general overviews see e.g. \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.

<!-- JN: maybe we can also add info about RMarkdown and similar, with R as an integrative tool for making scientific outputs -->

@@

<!-- MH: I went for a structure where in the first paragraph I briefly introduce the topic and then list the most common packages. -->
<!-- JN: I like this idea -->

@@

<!-- JN: maybe we should also add an introduction to R packages here-->
<!-- for example, CRAN vs GitHub vs other -->
<!-- who creates R packages? -->
<!-- who maintains them and what is an archived package -->
<!-- resources on how to create your own packages - https://r-pkgs.org/ -->
<!-- comment on never-ending changes in r world -->

@@

The use of landscape metrics was heavily facilitated by the introduction of the *FRAGSTATS* software \cite{McGarigal2012}, firstly published in 1995 \cite{Kupfer2012,Gustafson2019}.
However, *FRAGSTATS* is not not open-source software (however it is free to download at the developers' homepage), which restricts transparency, reproducibility, and extensibility.
To provide a truly open-source solution to calculate landscape metrics, recently the *landscapemetrics* package \cite{Hesselbarth2019a} was developed.
The package allows to calculate the most widely used landscape metrics in a transparent and reproducible workflow within the *R* environment. 
Additionally, the *belg* packages allows to calculate the Boltzmann entropy of a landscape gradient \cite{Nowosad2019a}.
<!-- JN: two notes: -->
<!-- 1. I am working on a paper supporting the belg package - I plan to submit it in less than a month, so hopefully we could update the reference here -->
<!-- 2. I also submitted a paper yesterday (2020-07-17) explaining a new package motif, which is aimed on calculations based on spatial signatures. I plan to add a paragraph about this package someplace in this paper. -->

@@

Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of *R* packages used for SDMs. 
Popular approaches and packages include generalized linear models using, e.g., the *stats* package; generalized additive models using, e.g., the *mgcv* or *lme4* package; classification and regression trees (CART) using, e.g., the *rpart*, *randomForest* or *ranger* package or multivariate data analysis using , e.g., the *ade4* or *vegan* package. 
Of course, also packages specifically designed for SDM exists. 
This includes the *dismo*, *sdm*, *ecospat*, *biomod2* and *PresenceAbsence* packages.
<!-- JN: I think all of the above packages should be cited -->

@@

Currently, to our knowledge, only one *R* packages exists that allows the use of neutral landscapes. 
<!--JN: I am not sure if that's true as few NLMR functions are just wrappers around existing R functions from RandomFields.
Also, it is possible to simulate landscapes (gradient landscapes) using geostatistical packages, e.g., gstat.
-->
Using a wide collection of algorithms, the *NLMR* package \cite{Sciaini2018} provides framework to simulate neutral landscape rasters. 
<!-- JN: I would also add landscapetools here as it allows to categorize outputs of the NLMR package -->

>>>>

---
title: Open source landscape ecology tools
# subtitle: Do you have a subtitle? If so, write it here
# titlerunning: Open source landscape ecology tools
# authorrunning: Short form of author list if too long for running head
thanks: | 
    Grants or other notes about the article that should go on the front 
    page should be placed here. General acknowledgments should be placed at the
    end of the article.
authors: 
- name: Maximillian H.K. Hesselbarth
  address: Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
  email: mhk.hesselbarthh@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Author 3
  address: Adress author 3
  email: E-Mail author 3
- name: ...
  address: ...
  email: ...
keywords:
- keyword 1
- keyword 2
- keyword 3
- keyword 4
- ...
abstract: |
  max. (200 words) Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: | 
  \usepackage{caption}
  \usepackage{ltablex}
  \usepackage{lscape}
  \usepackage{multirow}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
editor_options: 
  chunk_output_type: console
---

@@

@@

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}. 
Open-source software includes all software which is released under a license that allows to freely use, modify and distribute the software \cite{St.Laurent2008}.
Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology \cite{Prlic2012,Powers2019}. 
Furthermore, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} and to customize existing software to their specific needs \cite{Steiniger2009}. 
Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost which allows to use the software without any costs in contrast to many quite expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.

One example of a successful open-source project is the *R* programming language and its *Comprehensive R Archive Network* (CRAN) for extensions (also called packages) \cite{RCoreTeam2019}. 
Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}. 
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006}.
This is documented by the growing *CRAN Task View: Analysis of Spatial Data* \cite{Bivand2019a} and *CRAN Task View: Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} that currently list almost 300 packages in total. 
Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely even much higher.
The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.

Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on tools implemented in the *R* programming language.
The growing body of *R* packages related to spatial data processing and analyzing results in a high capability of the *R* programming language for landscape ecology.
Because additionally to the handling of spatial data, also further task such as statistical modeling, the preparation of publication-ready figures and even the preparation of complete reports using *RMarkdown* \cite{Xie2018} can be done in the same working environment, the *R* programming language can be a very powerful tool.
Thus, in the first part we are give a comprehensive overview over existing *R* packages for landscape ecology. 
In the second part, we identify topics for which *R* packages are currently missing to the best of our knowledge. 
For more general overviews see e.g. \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.

@@

@@

<!-- JN: maybe we should also add an introduction to R packages here-->
<!-- for example, CRAN vs GitHub vs other -->
<!-- who creates R packages? -->
<!-- who maintains them and what is an archived package -->
<!-- resources on how to create your own packages - https://r-pkgs.org/ -->
<!-- comment on never-ending changes in r world -->
<!-- MH: I like this idea. I will add a short paragraph -->

@@

The use of landscape metrics was heavily facilitated by the introduction of the *FRAGSTATS* software \cite{McGarigal2012}, firstly published in 1995 \cite{Kupfer2012,Gustafson2019}.
However, *FRAGSTATS* is not not open-source software (however it is free to download at the developers' homepage), which restricts transparency, reproducibility, and extensibility.
To provide a truly open-source solution to calculate landscape metrics, recently the *landscapemetrics* package \cite{Hesselbarth2019a} was developed.
The package allows to calculate the most widely used landscape metrics in a transparent and reproducible workflow within the *R* environment. 
Additionally, the *belg* packages allows to calculate the Boltzmann entropy of a landscape gradient \cite{Nowosad2019a}.
<!-- JN: two notes: -->
<!-- 1. I am working on a paper supporting the belg package - I plan to submit it in less than a month, so hopefully we could update the reference here -->
<!-- 2. I also submitted a paper yesterday (2020-07-17) explaining a new package motif, which is aimed on calculations based on spatial signatures. I plan to add a paragraph about this package someplace in this paper. -->
<!-- MH: Sounds great -->

@@

Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of *R* packages used for SDMs. 
Popular approaches and packages include generalized linear models using, e.g., the *stats* package; generalized additive models using, e.g., the *mgcv* or *lme4* package; classification and regression trees (CART) using, e.g., the *rpart*, *randomForest* or *ranger* package or multivariate data analysis using , e.g., the *ade4* or *vegan* package. 
Of course, also packages specifically designed for SDM exists. 
This includes the *dismo*, *sdm*, *ecospat*, *biomod2* and *PresenceAbsence* packages.
<!-- JN: I think all of the above packages should be cited -->
<!-- MH: Yeah, will do. I was just to lazy  -->

@@

Currently, to our knowledge, only one *R* packages exists that allows the use of neutral landscapes. 
<!-- JN: I am not sure if that's true as few NLMR functions are just wrappers around existing R functions from RandomFields. Also, it is possible to simulate landscapes (gradient landscapes) using geostatistical packages, e.g., gstat. -->
<!-- MH: Well, but RandomFields and geostatistcal are not directly designed for NLM. Maybe need to relax the statement a little bit -->
Using a wide collection of algorithms, the *NLMR* package \cite{Sciaini2018} provides framework to simulate neutral landscape rasters. 
<!-- JN: I would also add landscapetools here as it allows to categorize outputs of the NLMR package -->
<!-- MH: I would rahter add a short sentence about landscapetools in the general spatial data section since the package can be used for all different kinds of tasks not only for NLMR -->

### 2020-07-20 paper/paper.tex
Replying to some comments (either by comments or changes in the text)

\institute{
        Maximillian H.K. Hesselbarth \at
     Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany \\
     \email{\href{mailto:maximilian.hesselbarth@uni-goettingen.de}{\nolinkurl{maximilian.hesselbarth@uni-goettingen.de}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Jakub Nowosad \at
     Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland \\
     \email{\href{mailto:nowosad.jakub@gmail.com}{\nolinkurl{nowosad.jakub@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Author 3 \at
     Adress author 3 \\
     \email{E-Mail author 3}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        \ldots{} \at
     \ldots{} \\
     \email{\ldots{}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

@@

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
Open-source software includes all software which is released under a license that allows to freely use, modify and distribute the software \cite{St.Laurent2008}.
Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology \cite{Prlic2012,Powers2019}.
Lastly, open-source software allows other scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012} and to customize existing software to their specific needs \cite{Steiniger2009}.

One example of a successful open-source project is the \emph{R} programming language and its \emph{Comprehensive R Archive Network} (CRAN) for extensions (also called packages) \cite{RCoreTeam2019}.
Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \emph{R} programming language \cite{Bivand2006}.
This is documented by the growing \emph{CRAN Task View: Analysis of Spatial Data} \cite{Bivand2019a} and \emph{CRAN Task View: Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} that currently list almost 300 packages in total.
The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.

Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on tools implemented in the \emph{R} programming.
Thus, in the first part we are give a comprehensive overview over existing \emph{R} packages for landscape ecology.
In the second part, we identify topics for which \emph{R} packages are currently missing to the best of our knowledge.
For more general overviews see e.g.~\cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.

>>>>

\institute{
        Maximillian H.K. Hesselbarth \at
     Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany \\
     \email{\href{mailto:mhk.hesselbarthh@gmail.com}{\nolinkurl{mhk.hesselbarthh@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Jakub Nowosad \at
     Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland \\
     \email{\href{mailto:nowosad.jakub@gmail.com}{\nolinkurl{nowosad.jakub@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Author 3 \at
     Adress author 3 \\
     \email{E-Mail author 3}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        \ldots{} \at
     \ldots{} \\
     \email{\ldots{}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

@@

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
Open-source software includes all software which is released under a license that allows to freely use, modify and distribute the software \cite{St.Laurent2008}.
Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology \cite{Prlic2012,Powers2019}.
Furthermore, open-source software allows other scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012} and to customize existing software to their specific needs \cite{Steiniger2009}.
Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost which allows to use the software without any costs in contrast to many quite expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.

One example of a successful open-source project is the \emph{R} programming language and its \emph{Comprehensive R Archive Network} (CRAN) for extensions (also called packages) \cite{RCoreTeam2019}.
Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the \emph{R} programming language \cite{Bivand2006}.
This is documented by the growing \emph{CRAN Task View: Analysis of Spatial Data} \cite{Bivand2019a} and \emph{CRAN Task View: Handling and Analyzing Spatio-Temporal Data} \cite{Pebesma2020} that currently list almost 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of \emph{R} packages related to spatial data is most likely even much higher.
The growing popularity of the \emph{R} programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.

Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on tools implemented in the \emph{R} programming language.
The growing body of \emph{R} packages related to spatial data processing and analyzing results in a high capability of the \emph{R} programming language for landscape ecology.
Because additionally to the handling of spatial data, also further task such as statistical modeling, the preparation of publication-ready figures and even the preparation of complete reports using \emph{RMarkdown} \cite{Xie2018} can be done in the same working environment, the \emph{R} programming language can be a very powerful tool.
Thus, in the first part we are give a comprehensive overview over existing \emph{R} packages for landscape ecology.
In the second part, we identify topics for which \emph{R} packages are currently missing to the best of our knowledge.
For more general overviews see e.g.~\cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
